---
title: "Practical Machine Learning - Prediction Assignment"
author: "Mnolia Chikanya"
date: "2017 M02 22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Summary 
This report uses machine learning algorithms to predict the manner in which the subjects performed weight lifting exercises. The data is collected from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. The outcome variable has five classes and the total number of predictors are 159.


## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

## Preparation

### Loading the libraries and setting seed for reproducibility

```{r, results='hide'}
library(caret)
library(rpart)
library(knitr)
library(randomForest)
library(ElemStatLearn)
set.seed(2020)
```

### Loading the data

```{r}
# set the URL for the download
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the datasets
training <- read.csv(url(UrlTrain))
testing  <- read.csv(url(UrlTest))

# create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
```

```{r}
dim(TestSet)
```

### Data Cleaning

First we need to get rid of the variables that have close to zero variance in both training and testing data. Then remove the columns with missing values to avoid issues in training models. We remove the ID variables as well.
```{r}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
```

```{r}
# remove variables that are mostly NA
AllNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
```

```{r}
# remove identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
```

##  Prediction Model Building
Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model, as described below.

### Random Forest

```{r}
# model fit
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
                          trControl=controlRF)
# prediction on Test dataset
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest
```

### Decision Trees

```{r}
# model fit
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree
```

### Generalized Boosted Model

```{r}
# model fit
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
# prediction on Test dataset
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM
```

## Conclusion

The accuracy of the 3 regression modeling methods above are:

a. Random Forest : 0.9963
b. Decision Tree : 0.7368
c. GBM : 0.9839

In that case, the Random Forest model will be applied to predict the 20 quiz results (testing dataset) as shown below.

### Prediction with the Testing Dataset
```{r}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```




